---
title: "Preleminary ST analysis"
author: "Konstantinos Charitidis"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: '3'
    df_print: paged
  bookdown::html_document2:
    toc: true
    toc_depth: 3
    toc_float:
      collapse: false
    number_sections: true
    df_print: paged
    theme: readable
    code_folding: hide
    self_contained: true
    highlight: tango
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE)
options(max.print = 200)

```

# 1. Introduction # 
The dataset used for this analysis involves spatial transcriptomic data taken from lupus erythematosus patients which have undergone Car-T cell therapy. The ST data were retrieved from a CosMx machine. As of the moment the provided files are: 
1. A raw expression matrix 
2. A metadata file 
3. Md5sum files 
4. Seurat objects for two slides
5. One TIFF file for 4 samples?

- ** Input files: ** 


```{r load_libs, echo = FALSE, eval = TRUE}

library(SingleCellExperiment)
library(Seurat)
library(semla)
library(ggplot2)
library(dplyr)
library(rjson)
library(RColorBrewer)
library(magrittr)
library(pheatmap) 
library(tidyverse)

```


