---
title: "Loading and preprocessing of the CART data"
author: "Konstantinos Charitidis"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: '3'
    df_print: paged
  bookdown::html_document2:
    toc: true
    toc_depth: 3
    toc_float:
      collapse: false
    number_sections: true
    df_print: paged
    theme: readable
    code_folding: hide
    self_contained: true
    highlight: tango
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE)
options(max.print = 200)

```

```{r Libraries, include = FALSE}

library(tidyverse)
library(Seurat)
library(SeuratObject)
library(tiff)
library(Matrix)
library(patchwork)
source(file = "My_Functions.R")

```

```{r Load the files}

# My_Seurat_objs <- Load_My_Rds("../Data/")
# Raw_Counts <- read_csv("../Data/ASCLST12502_rawcount_exprMatrix.csv")
# Metadata <- read_csv("../Data/ASCLST12502_metadata.csv")

# Metadata <- Correct_Dataframes(Metadata) %>% 
#   rename(Cell_ID = cell_id) %>% 
#   select(Cell_ID, fov, x_FOV_px, y_FOV_px, x_slide_mm, y_slide_mm, everything())
# 
# Cell_IDs <- Raw_Counts[1]
# 
# Raw_Counts <- Correct_Dataframes(Raw_Counts) 
# 
# Raw_Counts["Cell_ID"] <- Cell_IDs
# 
# Raw_Counts <- Raw_Counts %>% 
#   select(Cell_ID, everything())
# 
# write_csv(x = Raw_Counts, file = "../Data/ASCLST12502_rawcount_fixed.csv")
# write_csv(x = Metadata, file = "../Data/ASCLST12502_metadata_fixed.csv")

Raw_Counts <- read_csv("../Data/ASCLST12502_rawcount_fixed.csv")
Metadata <- read_csv("../Data/ASCLST12502_metadata_fixed.csv")

```

```{r Basic information on the pre-existing Seurat objects}

# ## Explore which files were loaded ## 
# names(My_Seurat_objs)
# 
# ## Select file to analyse ##
# SeO_1 <- My_Seurat_objs$SeO_1
# # SeO_1 <- UpdateSeuratObject(SeO_1)
# 
# ## Basic information on the slide ##
# dim(SeO_1) 
# Assays(SeO_1) 
# DefaultAssay(SeO_1)
# Reductions(SeO_1) 
# Images(SeO_1)  
# 
# ## Basic metadata information ##
# head(SeO_1@meta.data, 5)
# names(SeO_1@meta.data)
# Idents(SeO_1) %>% head()
# table(Idents(SeO_1)) 
# 
# ## Information on spatial locations ##
# cands <- grep("(x|y|px|mm)", colnames(SeO_1@meta.data), value = TRUE, ignore.case = TRUE)
# cands

```

```{r Basic quality control}

# cols <- intersect(c("nFeature_RNA", "nCount_RNA", "percent.mt"), colnames(SeO_1@meta.data))
# if (length(cols)) VlnPlot(SeO_1, features = cols, pt.size = 0)

```

```{r Assay level check}

# DefaultAssay(SeO_1) <- Assays(SeO_1)[1]
# dim(GetAssayData(SeO_1, slot = "counts"))
# head(VariableFeatures(SeO_1), 10)

```

```{r Visualisation}

# DimPlot(SeO_1, reduction = "Export_PCA.1_1")
# DimPlot(SeO_1, reduction = "Export_UMAP.1_1")
# # SpatialDimPlot(SeO_1, images = Images(SeO_1)[1])

```

```{r Troubleshoot the image error}

# packageVersion("Seurat"); packageVersion("SeuratObject")
# 
# img_nm <- Images(SeO_1)[1]
# img_nm
# class(SeO_1@images[[img_nm]])
# 
# coords <- tryCatch(GetTissueCoordinates(SeO_1, image = img_nm), error = function(e) NULL)
# 
# if (!is.null(coords) && nrow(coords)) {
#   SpatialDimPlot(SeO_1, images = img_nm, crop = FALSE, combine = FALSE)[[1]]
# } else {
#   message("No tissue coordinates found for: ", img_nm)
# }

```

This is printing out this error:
Error in `geom_spatial()`:
! Problem while converting geom to grob.
ℹ Error occurred in the 1st layer.
Caused by error in `unit()`:
! 'x' and 'units' must have length > 0
Run `rlang::last_trace()` to see where the error occurred.
Error during wrapup: no slot of name "misc" for this object of class "FOV"
Error: no more error handlers available (recursive errors?); invoking 'abort' restart

Even though there are spot coordinates no plot is printed which means there is a problem with plotting the background image. 

```{r Plot the cells on top of the background manually}

# ## Load the TIFF as an array ##
# bg <- readTIFF(bg_path, native = FALSE)
# 
# ## Get the dimentions for the plot ##
# W <- dim(bg)[1]
# H <- dim(bg)[2]
# 
# ## Convert to raster for the overlay ##
# bg <- as.raster(bg)
# 
# ## Get the metadata and the scaling for CosMx (0.18 pixels per mm) ##
# md <- SeO_1@meta.data
# mm_per_px <- 0.18
# 
# ## Transform the values ##
# df <- transform(md, x_px = x_slide_mm * mm_per_px,
#                     y_px = y_slide_mm * mm_per_px) %>% 
#   select(x_px, y_px, fov)
# 
# ## Get the ranges for the annotation ##
# xr <- range(df$x_px)
# yr <- range(df$y_px)
# 
# ## Visualise ##
# New_plot <- ggplot(df, aes(x = x_px, y = y_px)) +
#   annotation_raster(bg, xmin = xr[1], xmax = xr[2],
#                     ymin = yr[1], ymax = yr[2],
#                     interpolate = TRUE) +
#   geom_point(size = 0.000001, colour = "white") +
#   coord_fixed() +
#   theme_void()
# 
# ## Save ##
# # ggsave(filename = "Coords_with_background.png",
# #        plot = New_plot,
# #        path = "../Results/QC_Images/",
# #        width = W, height = H,
# #        dpi = 300, units = "px")

```

```{r Visualise just the FOVs}

# ggplot(df, aes(x = x_px, y = y_px, colour = factor(fov))) +
#   geom_point(size = 0.03, alpha = 0.9) +
#   coord_fixed() +
#   theme_void() +
#   guides(colour = guide_legend(override.aes = list(size = 2))) +
#   labs(colour = "FOV")

```

```{r Create a subset to speed things up}

Wanted_Areas <- c("1", "2", "3", "34", "35", "36", "49", "50", "51", "69", "70", "71")

Metadata_sub <- Metadata %>% 
  filter(fov %in% Wanted_Areas)

Raw_Counts_sub <- Raw_Counts %>%
  filter(Cell_ID %in% Metadata_sub$Cell_ID)

keep <- c("Metadata_sub", "Raw_Counts_sub")
rm(list = setdiff(ls(envir = .GlobalEnv), keep), envir = .GlobalEnv)
gc()

# write_csv(x = Raw_Counts_sub, file = "../Data/Edited_data/Raw_Counts_sub.csv")
# write_csv(x = Metadata_sub, file = "../Data/Edited_data/Metadata_sub.csv")

```


```{r Create a Seurat objet with the raw counts}

## Load the files ##
# Raw_Counts_sub <- read_csv("../Data/Edited_data/Raw_Counts_sub.csv")
# Metadata_sub <- read_csv("../Data/Edited_data/Metadata_sub.csv") 

## Make sure the files contain the same cells ##
Common_cells <- intersect(Raw_Counts_sub$Cell_ID, Metadata_sub$Cell_ID)

Raw_Counts_sub <- Raw_Counts_sub %>%
  filter(Cell_ID %in% Common_cells) 

Metadata_sub <- Metadata_sub %>% 
  filter(Cell_ID %in% Common_cells) 

# Cell_IDs <- as.character(Raw_Counts_sub$Cell_ID)
# Gene_IDs <- names(Raw_Counts_sub)[-1]

Raw_Counts_sub_tidy <- Raw_Counts_sub %>%
  mutate(across(-Cell_ID, ~ suppressWarnings(as.numeric(.)))) %>%
  replace(is.na(.), 0) %>% 
  select(Cell_ID, everything())

Raw_Counts_matrix <- Raw_Counts_sub_tidy %>% 
  column_to_rownames(var = "Cell_ID") %>% 
  as.matrix() %>% 
  t()

Raw_Counts_sparse <- Matrix(Raw_Counts_matrix, sparse = TRUE)

dim(Raw_Counts_sparse)

```

```{r Create the Seurat Object}

Tissue_bg <- readTIFF(source = "../Data/MCF-ASCIST-1250-Slide-CART (1979 8769 6063 4322).tiff", native = FALSE)
Tissue_raster <- as.raster(Tissue_bg)

SeO_1_raw <- CreateSeuratObject(counts = Raw_Counts_sparse,
                                meta.data = Metadata_sub, 
                                project = "CosMx_CART", 
                                assay = "RNA", 
                                min.cells = 0, 
                                min.features = 0)

SeO_1_raw@misc$background <- Tissue_raster

```


```{r Seurat quality control}
VlnPlot(SeO_1_raw, features = c("nFeature_RNA", "nCount_RNA"), ncol = 2)

## Get total counts for each cell ##
SeO_1_raw$Total_Counts  <- Matrix::colSums(GetAssayData(SeO_1_raw, "RNA"))
## Get number of features/genes for each cell ##
SeO_1_raw$Num_Features  <- Matrix::colSums(GetAssayData(SeO_1_raw, "RNA") > 0)

## Get a visual understanding of their distributions ##
VlnPlot(SeO_1_raw, features = c("Total_Counts","Num_Features"), log = TRUE)

ggplot(SeO_1_raw@meta.data, aes(Total_Counts, Num_Features)) + 
  geom_point(alpha = 0.3)

```


```{r Filtering based on genes and cells}

## Filter based on geometric means of the areas ##
Geom_mean_area <- SeO_1_raw@meta.data %>%
  ## Get the cell area sizes ##
  transmute(Area_um2 = .data$Area.um2) %>%
  filter(is.finite(Area_um2), Area_um2 > 0) %>%
  ## Calculate the geometric mean of the areas ##
  summarize(Geom_mean = exp(mean(log(Area_um2), 
                          na.rm = TRUE)), 
            .groups = "drop") %>%
  ## Get the Geometric means ##
  pull(Geom_mean)

## Set the threshold for the cell filtering ##
Threshold_area <- 5 * Geom_mean_area

Area_um2_values <- SeO_1_raw@meta.data$Area.um2

## Filter cells with less than 30 transcripts and an area larger than the threshold ##
keep_cells <- with(SeO_1_raw@meta.data, Total_Counts >= 30 & is.finite(Area_um2_values) & Area_um2_values <= Threshold_area)
keep_cells <- rownames(SeO_1_raw@meta.data)[keep_cells]

## Subset based on the cell filtering and gene counts ##
SeO_1_filtered <- subset(SeO_1_raw, cells = keep_cells)

## Select genes that are transcribed in at least 1 cell ##
## GetAssayData retrieves the counts (Do this in the filtered object )##
keep_genes <- GetAssayData(SeO_1_filtered, assay = "RNA", layer = "counts") %>%
  ## rowSums checks if they are transcribed ##
  { rowSums(. > 0) } %>% 
  ## Conversion of the vector to a tibble/df ##
  enframe(name = "Gene", value = "nonzero") %>%
  ## Keeps the expressed genes #
  filter(nonzero > 0) %>%
  ## Get the gene names only to work with subset() ##
  pull(Gene)

SeO_1_filtered <- subset(SeO_1_filtered, features = keep_genes)

cat(sprintf("Kept %d cells, %d genes\nArea threshold used: %.3f\n", 
            ncol(SeO_1_filtered), nrow(SeO_1_filtered), Threshold_area))

```

```{r Visual QC}

Counts_filtered <- GetAssayData(SeO_1_filtered, layer = "counts")
Mitochondrial_gene_num <- grep("^MT-", rownames(Counts_filtered))

SeO_1_filtered$percent.mt <- Matrix::colSums(Counts_filtered[Mitochondrial_gene_num, , drop = FALSE]) /
  Matrix::colSums(Counts_filtered) * 100

paste0("Mitochondrial genes detected: ", length(Mitochondrial_gene_num))

SeO_1_filtered$Total_Counts <- Matrix::colSums(Counts_filtered)
SeO_1_filtered$Num_Features <- Matrix::colSums(Counts_filtered > 0)

VlnPlot(SeO_1_filtered, features = c("Total_Counts", "Num_Features", "percent.mt"), 
        pt.size = 0, ncol = 3)

FeatureScatter(SeO_1_filtered, feature1 = "Total_Counts", feature2 = "Num_Features")
FeatureScatter(SeO_1_filtered, feature1 = "Total_Counts", feature2 = "percent.mt")

SeO_1_filtered@meta.data %>%
  group_by(Area) %>% ## Area most likely contains the FOV information, needs adjustment ##
  summarise(Cells = n(), Median_Counts = median(Total_Counts, na.rm = TRUE),
            Median_Features = median(Num_Features, na.rm = TRUE),
            Median_MT = if ("percent.mt" %in% names(cur_data_all())) 
              median(percent.mt, na.rm = TRUE) else NA_real_) %>%
  arrange(Median_Counts) %>%
  print(n = 40)

```

```{r Background and negative probe quality control}

SeO_1_filtered@meta.data <- SeO_1_filtered@meta.data %>%
  ## Make sure all are numeric ##
  mutate(across(c(propNegative, nCount_negprobes, nCount_falsecode, Total_Counts),
                ~ suppressWarnings(as.numeric(.))),
        ## Compute background ratios ##
        prop_negprobes_cnt = nCount_negprobes / Total_Counts,
        prop_falsecode_cnt = nCount_falsecode / Total_Counts,
        ## Final background metric ##
        propNegative = coalesce(propNegative, prop_negprobes_cnt))

bg_plots <- list()

if ("propNegative" %in% colnames(SeO_1_filtered@meta.data)) {
  bg_plots[["propNeg"]] <- VlnPlot(SeO_1_filtered, features = "propNegative",
                                   group.by = "Area", pt.size = 0) + 
    ggtitle("Proportion of Negative Probe Signal by FOV")
}

if ("prop_falsecode_cnt" %in% colnames(SeO_1_filtered@meta.data)) {
  bg_plots[["false"]] <- VlnPlot(SeO_1_filtered, features = "prop_falsecode_cnt",
                                 group.by = "Area", pt.size = 0) + 
    ggtitle("Proportion of False Code Signal by FOV")
}

wrap_plots(bg_plots)

SeO_1_filtered@meta.data %>%
  group_by(Area) %>%
  summarise(n_cells = n(), 
            median_neg = median(propNegative, na.rm = TRUE),
            median_false = median(prop_falsecode_cnt, na.rm = TRUE),
            median_counts = median(Total_Counts, na.rm = TRUE)) %>%
  arrange(desc(median_neg))

SeO_1_filtered <- subset(SeO_1_filtered, subset = (is.na(propNegative) | propNegative < 0.05) &
                           (is.na(prop_falsecode_cnt) | prop_falsecode_cnt < 0.05))

summary(SeO_1_filtered@meta.data$propNegative)
summary(SeO_1_filtered@meta.data$prop_falsecode_cnt)

thr_neg <- quantile(SeO_1_filtered@meta.data$propNegative, 0.99, na.rm = TRUE)
thr_fc <- quantile(SeO_1_filtered@meta.data$prop_falsecode_cnt, 0.99, na.rm = TRUE)
cat("Suggested thresholds -> propNegative:", thr_neg, " prop_falsecode_cnt:", thr_fc, "\n")

if (!is.finite(thr_neg)) thr_neg <- 0.05
if (!is.finite(thr_fc))  thr_fc  <- 0.05

keep_bg <- (is.na(SeO_1_filtered@meta.data$propNegative) | 
              SeO_1_filtered@meta.data$propNegative < thr_neg) &
           (is.na(SeO_1_filtered@meta.data$prop_falsecode_cnt) | 
              SeO_1_filtered@meta.data$prop_falsecode_cnt < thr_fc)

cat("Will keep", sum(keep_bg, na.rm = TRUE), "of", ncol(SeO_1_filtered), "cells\n")
stopifnot(sum(keep_bg, na.rm = TRUE) > 0)

cells_keep <- colnames(SeO_1_filtered)[keep_bg]
SeO_1_filtered <- subset(SeO_1_filtered, cells = cells_keep)

## Re filter the genes in case some of the removed resulted in counts of 0 ##
keep_genes <- GetAssayData(SeO_1_filtered, assay = "RNA", layer = "counts") %>%
  { rowSums(. > 0) } %>% 
  enframe(name = "Gene", value = "nonzero") %>%
  filter(nonzero > 0) %>%
  pull(Gene)

SeO_1_filtered <- subset(SeO_1_filtered, features = keep_genes)

VlnPlot(SeO_1_filtered, features = intersect(c("propNegative", "prop_falsecode_cnt"), 
                                             colnames(SeO_1_filtered@meta.data)), pt.size = 0)


## Save Rds object to continue from here next session ##
# SaveSeuratRds(object = SeO_1_filtered, file = "../Data/Edited_data/SeO_1_After_bg_filtering.Rds")

```

```{r Filtering based on spatial}

Metadata_filtered <- SeO_1_filtered@meta.data

xr <- range(Metadata_filtered$x_slide_mm, na.rm = TRUE)
yr <- range(Metadata_filtered$y_slide_mm, na.rm = TRUE)

summary(SeO_1_filtered@meta.data$x_slide_mm)
summary(SeO_1_filtered@meta.data$y_slide_mm)
length(unique(SeO_1_filtered@meta.data$Area))

SeO_1_filtered@meta.data %>%
  group_by(Area) %>%
  summarise(xmin = min(x_slide_mm, na.rm = TRUE),
            xmax = max(x_slide_mm, na.rm = TRUE),
            ymin = min(y_slide_mm, na.rm = TRUE),
            ymax = max(y_slide_mm, na.rm = TRUE))

## Total counts in space ##
ggplot(Metadata_filtered, aes(x = x_slide_mm, y = y_slide_mm, color = Total_Counts)) +
  geom_point(size = 0.1, alpha = 0.8) +
  coord_fixed(xlim = xr, ylim = yr) +
  scale_color_viridis_c() +
  theme_void() + 
  ggtitle("Total counts (per cell)")

## Number of features in space ##
ggplot(Metadata_filtered, aes(x = x_slide_mm, y = y_slide_mm, color = Num_Features)) +
  geom_point(size = 0.1, alpha = 0.8) +
  coord_fixed(xlim = xr, ylim = yr) +
  scale_color_viridis_c() +
  theme_void() + 
  ggtitle("Features detected (per cell)")

## Background fractions in space (if present) ##
if ("propNegative" %in% names(Metadata_filtered)) {
  ggplot(Metadata_filtered, aes(x = x_slide_mm, y = y_slide_mm, color = propNegative)) +
    geom_point(size = 0.1, alpha = 0.8) +
    coord_fixed(xlim = xr, ylim = yr) +
    scale_color_viridis_c(limits = c(0, quantile(Metadata_filtered$propNegative, 0.99, na.rm = TRUE))) +
    theme_void() + 
    ggtitle("Background (propNegative)")
}

if ("prop_falsecode_cnt" %in% names(Metadata_filtered)) {
  ggplot(Metadata_filtered, aes(x = x_slide_mm, y = y_slide_mm, color = prop_falsecode_cnt)) +
    geom_point(size = 0.1, alpha = 0.8) +
    coord_fixed(xlim = xr, ylim = yr) +
    scale_color_viridis_c(limits = c(0, quantile(Metadata_filtered$prop_falsecode_cnt, 0.99, na.rm = TRUE))) +
    theme_void() + 
    ggtitle("Background (false code fraction)")
}

## Counts per FOV ##
ggplot(Metadata_filtered, aes(x = x_slide_mm, y = y_slide_mm, color = Total_Counts)) +
  geom_point(size = 0.07, alpha = 0.8) +
  # coord_fixed() +
  scale_color_viridis_c() +
  facet_wrap(~ Area, scales = "free") +
  theme_void() + 
  ggtitle("Total counts per FOV")

## Background per FOV ##
if ("propNegative" %in% names(Metadata_filtered)) {
  ggplot(Metadata_filtered, aes(x = x_slide_mm, y = y_slide_mm, color = propNegative)) +
    geom_point(size = 0.07, alpha = 0.8) +
    coord_fixed() +
    scale_color_viridis_c() +
    facet_wrap(~ Area, scales = "free") +
    theme_void() + 
    ggtitle("propNegative per FOV")
}

```




